The Name of the Title Is Hope

### 2025-01-16 paper.tex
Incorporate feedback, add preliminary graphics.

\section{Introduction}
(Tell readers we are targeting WebGPU)
(Speed of light definition here)
\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures.
\end{itemize}

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier''). There is a fundamental reason why this primitive is expensive. Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups},\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.} which are then grouped into larger units called \emph{workgroups}.\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.} This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory;\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).} and all threads have access to high-bandwidth but higher-latency global memory.

While threads within the same workgroup can synchronize with each other using barrier primitives and can communicate through shared memory, a workgroup is a logical unit of execution, not a physical processor. Instead, workgroups are dynamically mapped and scheduled to a GPU's underlying \emph{multiprocessors}. This abstraction enables the kernel to execute on GPUs with varying hardware capabilities, such as different numbers of multiprocessors or multiprocessors with differing execution resources. A single multiprocessor can host multiple workgroups; we define the number of active subgroups on a multiprocessor as the \emph{subgroup occupancy}, and the number of active workgroups across the entire GPU as the \emph{workgroup occupancy}. Generally speaking, high subgroup occupancy is desirable because multiprocessors are specifically designed to context switch between subgroups, keeping execution units busy and masking memory and execution latency.

GPUs exhibit both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors. A multiprocessor schedules and executes threads by subgroup. While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD\@. However, this flexibility comes at the cost of subgroup divergence. When threads in a subgroup diverge due to conditional branching or other control flow, each unique path is executed sequentially, but at the full execution width, effectively rendering the operation serial and reducing parallel efficiency.\footnote{The scheduling and execution width of the underlying hardware does not necessarily align with the subgroup width.}

Most relevant to this work is the scheduling behavior of the GPU\@. Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution. Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model. Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch. At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues. As we will discuss in the following section, these factors have profound implications for kernel performance and correctness.

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. \john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic. I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious! explain why}. The introduction of shared memory \john{I think it's not just this, it's also hierarchical compute, grid$\rightarrow$workgroup$\rightarrow$thread. The important thing here is not just saying ``this is the contemporary architecture'' but also WHY it's the contemporary architecture. I like how you are focusing on the tile dependencies as the important aspect.} and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies. \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory. However, as a natural result of this partitioning, a serial dependency is created between the work tiles. In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system. So the first goal is to find an algorithmic formulation that requires a minimum memory volume. Then given the above, and given a particular algorithmic formulation, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth. Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}. \john{For large scans, this stage is an insignificant amount of work compared to the first and third.}
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. \john{Language is a little sloppy here, since the root is not propagated back to each workgroup; instead each workgroup gets one element of the root, yes? I like that you are using and distinguishing root and spine; I just want them to be precise.}
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. An input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\john{I liked what you said above (Section~\ref{sec:the-scan-primitive}) about how each of these formulations ``navigate this inter-workgroup dependency''. Make that explicit in this subsubsection, and each of the below!}

\subsubsection{Reduce-then-Scan}
Due to the extreme latency \john{I don't think this is a latency problem. GPUs tolerate long latency well. This is a bandwidth problem. We don't have enough bandwidth to keep the compute units fed.} of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. \john{I would try to use the word ``tradeoff'' here, because you'll use it again. In this case, we profit from trading additional computation for less memory bandwidth. We want to come back to ``scan is memory-bound, so reducing memory traffic, even at the cost of additional computation, is profitable''.} This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. \john{The important part here IMHO is changing the problem size in the second step from ``variable sized, and proportional to the problem size'' (which might require recursion) to ``constant sized, and thus guaranteed to avoid recursion''. I would identify the former as an important problem to be solved because of the potential need for further recursion, and the latter as an elegant solution to it.} By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes \john{I think you want to say, explicitly, the work that each workgroup actually does, and note what part of that is latency-sensitive (since, as you say below, the latency of the sequential dependency is super important for overall scan performance.} its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup. Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory. Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$. \john{Somewhere in here you want to put in at least a hint of ``forward progress guarantee''. You don't have to use those words. But when you get to FPG, you want to point back and say ``this formulation depends on \ldots''.}

Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by sequential dependencies. Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}—a serial traversal backward along the spine to resolve dependencies. The work performed during this phase is constant-bounded to the workgroup occupancy $o$, as redundant operations are avoided by updating the workgroup's reduction from local to inclusive once the lookback is completed. Although this increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing ensures that memory accesses to the spine are roughly contiguous within the workgroup occupancy block, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces a negligible latency overhead. With a total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance.

@@

  $iteration\_offset \gets 0$\;
  $stride \gets 1$\;

@@

      $spine\_index \gets ((thread\_id + iteration\_offset) * stride) - iteration\_offset$\;

      \If{$spine\_index < spine\_length$}{
        $x[spine\_index] \gets \text{subgroupInclusiveScan}(x[spine\_index])$\;
      }

@@

\subsection{Tradeoffs}

>>>>

\section{Introduction (WIP)}
(Tell readers we are targeting WebGPU)
(Speed of light definition here)
\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures.
\end{itemize}
\thomas{See comment in section 3 !}
Our method retains the best aspects of the \emph{Chained-Scan} architecture, while eliminating the need for forward progress guarantees. It is designed to operate on arbitrary hardware, support arbitrary monoids, process inputs of arbitrary size, and support arbitrary 32-bit and 64-bit data types.
\subsection{Goals}
This work is guided by two main goals:
\begin{enumerate}
  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG or support for 64-bit atomics.
  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
\end{enumerate}

\subsection{Non Goals}
Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models. As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance. Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in them.} placing a solution to subgroup divergence issues outside our scope. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

\subsection{Contributions}
\begin{enumerate}
    \item Decoupled-Fallback
    \item Split-Method
    \item Subgroup-Size Agnostic stuff
\end{enumerate}

\section{Background}
\subsection{Virtualization and Scheduling in the GPU programming model}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. (one sentence on thread hierarchy maybe). As comprehensive descriptions of the GPU programming model~\cite{} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubcription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.

\subsubsection{Inter-Workgroup Barriers}
Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\subsubsection{Fairness and Progress}
At the subgroup scheduling level, two key issues emerge: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.~\cite{10.1145/3022671.2984032, sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358, Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.

\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
Nvidia, by contrast uses fairness in line with our language. From Tesla architecture: \newline
"A scoreboard qualifies each warp for issue each cycle. The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue. Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
From DecoupledLookback: \newline
"Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
"Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
}

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-6774(86)90003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers. To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory. Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans is memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}. Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization. In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
\end{enumerate}

As each work tile depends on the reductions of previous tiles, the absence of an inter-workgroup barrier necessitates kernel launches to resolve inter-workgroup dependencies. This results in the three-phase construction of \emph{Scan-then-Propagate}, with each phase implemented as a separate kernel launch. This is further complicated in Sengupta et al.'s implementation, where arbitrarily large inputs are processed through recursion. As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase scan structure must be recursively applied until the spine size fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like \emph{scan} can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies. However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merill and Grimshaw~\cite{} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two. In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup. This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion. Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$. Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{} was the first to apply more advanced inter-workgroup synchronization techniques to scan. Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches. Since multiprocessors are not oversubscribed, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching. Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads. \thomas{This also requires FPG, but not ideal to unpack and mention here . . .}

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and status flags to ensure coherent views of data. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and incorporate the reductions of all preceding tiles. In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the progress of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of its predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.

\subsubsection{Chained-Scan, Decoupled-Lookback}
Building on \emph{Stream-Scan}, Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on progress guarantees, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces a negligible latency overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.

@@

  $stride \gets 1$\;
  $top\_offset \gets 0$\;

@@

      $step \gets spine\_length / stride$

      \If{$thread\_id < step$}{
        $temp \gets \text{subgroupInclusiveScan}(x[thread\_id + top\_offset])$\;
        $x[thread\_id + top\_offset] \gets temp$\;
        \If{$\text{\normalfont highestRankLane}(thread\_id)$}{
          $x[(thread\_id / S) + step + top\_offset] \gets temp$;
        }
      }

@@

\subsection{Trade-offs}
